Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2022-11-11 pkg/glmertree/paper/glmm-tree/treatment_interactions.Rnw
moved first paper to glmm-tree folder
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3235 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b616a7442d9e5ce4d2444cd3cb972adaa86cf16e

\section{Abstract}
Identification of subgroups of patients for whom treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research. Therefore, we propose the generalized linear mixed-effects model tree (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters. In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than linear-model based recursive partitioning and mixed-effects regression trees. Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effects, on average. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods. Tree-based methods, also known as recursive partitioning methods, split observations repeatedly into groups so that they become increasingly similar with respect to the outcome within each group. Several tree-based methods take the mean of a continuous dependent variable or the majority class of a categorical dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}. Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM approach of \citeA{Ciam91} is the earliest example.

Due to the recursive nature of the splitting, the rectangular regions of a recursive partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}. The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety. The resulting tree has a depth of two, as the longest path travels along two splits. Each of the splits in the tree is defined by a splitting variable and value. The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of 8, yielding two rectangular regions, represented by node~2 and node~3. As the observations in node~2 are not further split, node~2 is a terminal node. Node~3 is an inner node, as the observations in this node are further split into nodes~4 and~5, based on the anxiety variable.

@@

\begin{figure}[tb!]
\setkeys{Gin}{width=0.5\textwidth}
\centering
\subfloat[Rectangular areas]{
<<example_decision_tree2, fig = TRUE>>=
options(prompt = " ", continue = " ")
plot(c(0, 18), c(0, 18), type = "n", xlab = "duration", ylab = "anxiety")
abline(v = 8, h = NA)
lines(x = c(-1, 8), y = c(10,10))
text(label = "node 3", x = 3.5, y = 4.5)
text(label = "node 4", x = 3.5, y = 14)
text(label = "node 5", x = 13.5, y = 9)
@
}
\subfloat[Decision tree]{
<<example-decision-tree, fig=TRUE>>=
fig1 <- party(
  partynode(1L,
    split = partysplit(1L, breaks = 8),
      kids = list(
        partynode(3L, 
          split = partysplit(2L, breaks = 10),
          kids = list(
            partynode(4L, info = ""),
            partynode(5L, info = ""))),
  partynode(2L, info = ""))),
  data.frame(duration = numeric(0), anxiety = numeric(0))
)
plot(fig1)
@
}
\caption{\small Example of a recursive partition. In the left panel, the partition is depicted as a set of rectangular areas. In the right panel, the same partition is depicted as a decision tree.}
\label{fig:exampletree}
\end{figure}

@@

The remainder of this paper is structured into four sections: In the first section, we introduce the GLMM tree algorithm using an artificial motivating dataset with treatment-subgroup interactions. In the second section, we compare the performance of GLMM trees with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions. In the third section, we apply the GLMM tree algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. In the fourth and last section we summarize the results and discuss limitations and directions for future research. A tutorial on how to fit GLMM trees using the R package \verb|glmertree| is included as supplementary material. In the tutorial, the artificial motivating dataset is used, allowing users to recreate the trees and models to be fitted in the next section.

@@

\begin{table}[h!]
\small
\caption{\small Summary statistics for partitioning and outcome variables in the artificial motivating dataset}
\begin{tabular}{lcccc}
\thickline
 & min & max & mean & sd\\
\hline
depression & 3 & 16 & 9.12 & 2.66 \\
age & 18 & 69 & 45 & 9.56 \\
anxeity & 3 & 18 &  10.26 & 3.05 \\
duration & 1 & 17 & 6.97 & 2.90 \\
\hline
\end{tabular}
\label{tab:sum_stats}
\end{table}

@@

\begin{figure}[tb!]
\centering
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{\small Example of a global GLM for treatment outcomes, based on the artificial motivating dataset ($N=150$). The y-axis represent treatment outcome (post-treatment depression). The dot for Treatment~1 represents the first and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

\begin{figure}[tb!]
\centering
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{\small GLM tree grown on the artificial motivating dataset. Three additional covariates (pre-treatment anxiety, duration and age) were used as potential splitting variables. The y-axes in the terminal nodes represent the treatment outcome (post-treatment depression severity).}
\label{fig:example_mobtree}
\end{figure}

The partition resulting from application of MOB can be depicted as a decision tree. If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%

@@

Although the random-effects part of the GLMM in Equation~\ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part $x_{i}^{\top}\beta$ may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation~\ref{eq:fixedeffects_MOB} while still adjusting for random effects: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
In the GLMM tree model, the fixed effects $\beta_j$ are local parameters, their value depending on subgroup $j$, but the random effects $b$ are global. To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects parameters are estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random-effects parameters.

\begin{figure}[tb!]
\centering
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{\small GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months) and age were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

  \item[Step 2:] Fit the mixed-effects model $g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b$ with subgroups $j(r)$ from the GLM tree estimated in Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.

@@

\begin{enumerate} 
\item Sample size: $N=200$, $N=500$, $N=1000$.
\item Number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
\item Intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Number of clusters: $M=5$, $M=10$, $M=25$.
\item Population standard deviation (SD) of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
\end{enumerate}

Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. Means for other potential partitioning covariates were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$.

@@

R \cite{R} was used for data generation and analyses. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15, pkg:partykit}) was employed for estimating LM trees, using the \verb|lmtree| function. For estimation of LMM trees the \verb|lmertree| function of the \verb|glmertree| package (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge) was used. The significance level $\alpha$ for the parameter instability tests was set to .05 for all trees, with a Bonferroni correction applied for multiple testing. The minimum number of observations per node in trees was set to 20 and maximum tree depth was set to three, thus limiting the number of terminal nodes to eight in every tree.

@@

\begin{figure}[tb!]
\centering
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
partynode(1L,
split = partysplit(2L, breaks = 30),
kids = list(
partynode(2L,
split = partysplit(1L, breaks = 17),
kids = list(
partynode(3L, info = c(
expression(''),
expression(beta[j0] == '17.5'),
expression(''),
expression(beta[j1] == -'5.0'),
expression(''),
expression(d[j] == -'1.0')
)),
partynode(4L, info = c(
expression(''),
expression(beta[j0] == '30.0'),
expression(''),
expression(beta[j1] == '0.0'),
expression(''),
expression(d[j] == '0.0')
)))),
partynode(5L,
split = partysplit(3L, breaks = 63),
kids = list(
partynode(6L, info = c(
expression(''),
expression(beta[j0] == '30.0'),
expression(''),
expression(beta[j1] == '0.0'),
expression(''),
expression(d[j] == '0.0')
)),
partynode(7L, info = c(
expression(''),
expression(beta[j0] == '42.5'),
expression(''),
expression(beta[j1] == '5.0'),
expression(''),
expression(d[j] == '1.0')
)))))),
data.frame(U1 = numeric(0), U2 = numeric(0), U5 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{\small Data-generating model for treatment-subgroup interactions. Parameter $d_j$ denotes the node-specific standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

For generating datasets with treatment-subgroup interactions, we used a design from \citeA{DussyMech14} which is depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows four subgroups, characterized by values of the partitioning variables $U_2$, and $U_1$ or $U_5$. Two of the subgroups have mean differences in treatment outcome, indicated by a non-zero value of $\beta_{j1}$, and two subgroups do not have mean differences in treatment outcome, indicated by a $\beta_{j1}$ value of 0.

@@

\begin{figure}[tb!]
\centering
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
treespecs.long2$N <- factor(as.numeric(as.character(treespecs.long2$N)), ordered = TRUE)
treespecs.long2$sigmab <- factor(as.numeric(as.character(treespecs.long2$sigmab)), ordered = TRUE)
levels(treespecs.long2$treesize.ind)[levels(treespecs.long2$treesize.ind)=="GLM tree"] <- "LM tree"
levels(treespecs.long2$treesize.ind)[levels(treespecs.long2$treesize.ind)=="GLMM tree"] <- "LMM tree"
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN = mean, 
                     data = treespecs.long2)
levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "bi and splitting U correlated"] <- "b correlated with splitting U"
levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "bi and non-splitting U correlated"] <- "b correlated with non-splitting U"
levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "uncorrelated"] <- "b and U uncorrelated"
xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups = treesize.ind, type = "b", 
       ylab = "tree size", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), abline = c(7,0),
       auto.key = list(space = "top", columns = 2, title = " ", cex.title = 1, lines = TRUE, points = TRUE))
@
\caption{\small Average tree size of LM and LMM trees. Tree size is defined as the total number of nodes in a tree. Rows represent the correlation between the random intercepts and one of the partitioning variables, columns represent sample size. The horizontal line in each graph indicates the 'true' tree size (the total number of nodes in the tree used for generating the interactions).}
\label{fig:xyplot_treesize}
\end{figure}

@@

\begin{figure}[tb!]
\centering
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
levels(treespecs.long2$truetree.ind)[levels(treespecs.long2$truetree.ind)=="GLM tree"] <- "LM tree"
levels(treespecs.long2$truetree.ind)[levels(treespecs.long2$truetree.ind)=="GLMM tree"] <- "LMM tree"
aggdata.acc <- aggregate(formula = truetree.values ~ truetree.ind + N + sigmab + corUb, FUN = mean, data = treespecs.long2)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups = truetree.ind, type = "b", 
ylab = "tree accuracy", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), 
auto.key = list(space = "top", columns = 2, title = " ", cex.title = 1, lines = TRUE, points = TRUE)))
@
\caption{\small Tree accuracy of LM and LMM tree. Tree accuracy is defined as the proportion of datasets in which the true tree was accurately recovered. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$, columns represent sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[tb!]
\centering
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLM tree"] <- "LM tree"
levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLMM tree"] <- "LMM tree"
aggdata.cor <- aggregate(formula = correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN = mean, data = treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups = correlation.ind, 
type = "b", ylab = "correlation", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), 
auto.key = list(space = "top", columns = 3, title = " ", cex.title = 1, lines = TRUE, points = TRUE), ylim = c(0.3, 1.1)))
@
\caption{\small Average predictive accuracy of LM and LMM trees. Predictive accuracy is defined as the correlation between the true and predicted treatment-effect differences. Rows represent absolute treatment-effect differences in subgroups with treatment-effect differences, columns represent sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

Compared to LMMs with pre-specified interactions, LMM trees provided somewhat better accuracy, on average. LMM trees performed much better than LMMs when interactions were at least partly piecewise. However, the performance ofLMM trees deteriorated when the interactions were purely continuous functions of the predictor variables, in which case LMMs with pre-specified interactions performed very well. The performance of LMMs deteriorated with a larger number of predictor variables, whereas the performance of LMM trees was not affected by this, confirming our expectation that LMM trees are better suited for exploration than LMMs.

@@

>>>>

\section{Abstract}
Identification of subgroups of patients for whom treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available algorithms allow for taking into account clustered or nested dataset structures, which are particularly common in psychological research. Therefore, we propose the generalized linear mixed-effects model tree (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning to detect treatment-subgroup interactions, and a GLMM to estimate the random-effects parameters. In a simulation study, GLMM trees show higher accuracy in recovering treatment-subgroup interactions, higher predictive accuracy, and lower Type-II error rates than linear-model based recursive partitioning and mixed-effects regression trees. Also, GLMM trees show somewhat higher predictive accuracy than linear mixed-effects models with pre-specified interaction effects, on average. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM trees are a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, mixed-effects models, classification and regression trees

@@

Tree-based methods are such hypothesis-generating methods. Tree-based methods, also known as recursive partitioning methods, split observations repeatedly into groups so that they become increasingly similar with respect to the outcome within each group. Several tree-based methods take the mean of a continuous dependent variable or the majority class of a categorical dependent variable as the outcome, one of the earliest and most well-know examples being the classification and regression tree (CART) approach of \citeA{BreiyFrie84}. Other tree-based methods take the estimated parameters of a more complex model, of which the RECPAM (recursive partition and amalgamation) approach of \citeA{Ciam91} is the earliest example.

Due to the recursive nature of the splitting, the rectangular regions of a recursive partition can be graphically depicted as nodes in a decision tree, as shown in the artificial example in Figure~\ref{fig:exampletree}. The partition in Figure~\ref{fig:exampletree} is rather simple, based on the values of two predictor variables: duration and anxiety. The resulting tree has a depth of two, as the longest path travels along two splits. Each of the splits in the tree is defined by a splitting variable and value. The first split in the tree separates the observations into two subgroups, based on the duration variable and a splitting value of eight, yielding two rectangular regions, represented by node~2 and node~5. Node~2 is an inner node, as the observations in this node are further split into terminal nodes~3 and~4, based on the anxiety variable. The observations in node~5 are not further split and this is therefore a terminal node.

@@

\begin{figure}[tb!]
\setkeys{Gin}{width=0.5\textwidth}
\centering
\subfloat[Rectangular areas]{
<<example_decision_tree2, fig = TRUE>>=
options(prompt = " ", continue = " ")
plot(c(0, 18), c(0, 18), type = "n", xlab = "duration", ylab = "anxiety")
abline(v = 8, h = NA)
lines(x = c(-1, 8), y = c(10,10))
text(label = "node 3", x = 3.5, y = 4.5)
text(label = "node 4", x = 3.5, y = 14)
text(label = "node 5", x = 13.5, y = 9)
@
}
\subfloat[Decision tree]{
<<example-decision-tree, fig=TRUE>>=
fig1 <- party(
  partynode(1L,
    split = partysplit(1L, breaks = 8),
      kids = list(
        partynode(3L, 
          split = partysplit(2L, breaks = 10),
          kids = list(
            partynode(4L, info = ""),
            partynode(5L, info = ""))),
  partynode(2L, info = ""))),
  data.frame(duration = numeric(0), anxiety = numeric(0))
)
plot(fig1)
@
}
\caption{\small Example recursive partition defined by two partitioning variables (duration and anxiety). In the left panel, the partition is depicted as a set of rectangular areas. In the right panel, the same partition is depicted as a tree.}
\label{fig:exampletree}
\end{figure}

@@

The remainder of this paper is structured into four sections: In the first section, we introduce the GLMM tree algorithm using an artificial motivating dataset with treatment-subgroup interactions. In the second section, we compare the performance of GLMM trees with that of three other methods: MOB trees without random effects, mixed-effects regression trees (MERTs) and linear mixed-effects models with pre-specified interactions. In the third section, we apply the GLMM tree algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapy for depression. In the fourth and last section we summarize the results and discuss limitations and directions for future research. In the Appendix, we provide a glossary explaining abbreviations and mathematical notation used in the current paper. Finally, a tutorial on how to fit GLMM trees using the R package \verb|glmertree| is included as supplementary material. In the tutorial, the artificial motivating dataset is used, allowing users to recreate the trees and models to be fitted in the next section.

@@

\begin{table}[h!]
\small
\caption{\small Summary statistics for partitioning and outcome variables in the artificial motivating dataset}
\begin{tabular}{lcccc}
\thickline
 & min & max & $M$ & $SD$\\
\hline
depression & 3 & 16 & 9.12 & 2.66 \\
age & 18 & 69 & 45.00 & 9.56 \\
anxeity & 3 & 18 &  10.26 & 3.05 \\
duration & 1 & 17 & 6.97 & 2.90 \\
\hline
\end{tabular}
\label{tab:sum_stats}
\end{table}

@@

\begin{figure}[tb!]
\centering
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{\small Example of a globally estimated GLM based on the artificial motivating dataset ($N=150$). The $x$-axis represents treatment group, the $y$-axis represents treatment outcome (post-treatment depression). The dot for Treatment~1 represents the first element of the fixed-effects coefficient vector $\beta$, the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

\begin{figure}[tb!]
\centering
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{\small GLM tree grown on the artificial motivating dataset. The $x$-axes in the terminal nodes represent the treatment group, the $y$-axes represent the treatment outcome (post-treatment depression). Three additional covariates (pre-treatment anxiety, duration and age) were used as potential splitting variables, of which two (duration and age) were selected.}
\label{fig:example_mobtree}
\end{figure}

The partition resulting from application of MOB can be depicted as a decision tree. If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node:
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%

@@

Although the random-effects part of the GLMM in Equation~\ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part $x_{i}^{\top}\beta$ may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation~\ref{eq:fixedeffects_MOB} while still adjusting for random effects: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
In the GLMM tree model, the fixed effects $\beta_j$ are local parameters, their value depending on terminal node $j$, but the random effects $b$ are global. To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects parameters are estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random-effects parameters.

\begin{figure}[tb!]
\centering
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{\small GLMM tree of the motivating example dataset. The $x$-axes represent the treatment group, the $y$-axes represent the treatment outcome (post-treatment depression). Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months) and age were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

  \item[Step 2:] Fit the mixed-effects model $g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b$ with terminal node $j(r)$ from the GLM tree estimated in Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.

@@

\begin{enumerate} 
\item Sample size: $N=200$, $N=500$, $N=1000$.
\item Number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
\item Intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Number of clusters: $M=5$, $M=10$, $M=25$.
\item Population standard deviation (SD) of the normal distribution from which the cluster-specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates ($r = .42$).  
\end{enumerate}

Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. The means of the other potential partitioning covariates ($U_3$ and, depending on the value of $K$, also $U_6$ through $U_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$.

@@

R \cite{R} was used for data generation and analyses. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15, pkg:partykit}) was employed for estimating LM trees, using the \verb|lmtree| function. For estimation of LMM trees the \verb|lmertree| function of the \verb|glmertree| package (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge) was used. The significance level $\alpha$ for the parameter instability tests was set to .05 for all trees, with a Bonferroni correction applied for multiple testing. The latter adjusts the $p$-values of the parameter stability tests, by multiplying these by the number of potential partitioning variables. The minimum number of observations per node in trees was set to 20 and maximum tree depth was set to three, thus limiting the number of terminal nodes to eight in every tree.

@@

\begin{figure}[tb!]
\centering
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
partynode(1L,
split = partysplit(2L, breaks = 30),
kids = list(
partynode(2L,
split = partysplit(1L, breaks = 17),
kids = list(
partynode(3L, info = c(
expression(''),
expression(beta[j0] == '17.5'),
expression(''),
expression(beta[j1] == -'5.0'),
expression(''),
expression(d[j] == -'1.0')
)),
partynode(4L, info = c(
expression(''),
expression(beta[j0] == '30.0'),
expression(''),
expression(beta[j1] == '0.0'),
expression(''),
expression(d[j] == '0.0')
)))),
partynode(5L,
split = partysplit(3L, breaks = 63),
kids = list(
partynode(6L, info = c(
expression(''),
expression(beta[j0] == '30.0'),
expression(''),
expression(beta[j1] == '0.0'),
expression(''),
expression(d[j] == '0.0')
)),
partynode(7L, info = c(
expression(''),
expression(beta[j0] == '42.5'),
expression(''),
expression(beta[j1] == '5.0'),
expression(''),
expression(d[j] == '1.0')
)))))),
data.frame(U1 = numeric(0), U2 = numeric(0), U5 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{\small Data-generating model for treatment-subgroup interactions. $U_1$, $U_2$ and $U_5$ represent the (true) partitioning variables; $\beta$ represents the (true) fixed effects, with $\beta_{j0}$ representing the intercept and $\beta_{j1}$ representing the slope; parameter $d_j$ denotes the node-specific standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

For generating datasets with treatment-subgroup interactions, we used a design from \citeA{DussyMech14} which is depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows four terminal subgroups, characterized by values of the partitioning variables $U_2$, and $U_1$ or $U_5$. Two of the subgroups have mean differences in treatment outcome, indicated by a non-zero value of $\beta_{j1}$, and two subgroups do not have mean differences in treatment outcome, indicated by a $\beta_{j1}$ value of 0.

@@

\begin{figure}[tb!]
\centering
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
treespecs.long2$N <- factor(as.numeric(as.character(treespecs.long2$N)), ordered = TRUE)
treespecs.long2$sigmab <- factor(as.numeric(as.character(treespecs.long2$sigmab)), ordered = TRUE)
levels(treespecs.long2$treesize.ind)[levels(treespecs.long2$treesize.ind)=="GLM tree"] <- "LM tree"
levels(treespecs.long2$treesize.ind)[levels(treespecs.long2$treesize.ind)=="GLMM tree"] <- "LMM tree"
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN = mean, 
                     data = treespecs.long2)
levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "bi and splitting U correlated"] <- "b correlated with splitting U"
levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "bi and non-splitting U correlated"] <- "b correlated with non-splitting U"
levels(aggdata.size$corUb)[levels(aggdata.size$corUb) == "uncorrelated"] <- "b and U uncorrelated"
xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups = treesize.ind, type = "b", 
       ylab = "tree size", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), abline = c(7,0),
       auto.key = list(space = "top", columns = 2, title = " ", cex.title = 1, lines = TRUE, points = TRUE))
@
\caption{\small Average tree size of LM and LMM trees. The $y$-axes represent the average number of nodes in trees; $x$-axes represent the population standard deviation of the normal distribution from which the cluster-specific intercepts were drawn; rows represent the correlation between the random intercepts and one of the partitioning variables; columns represent sample size. The horizontal line in each graph indicates the 'true' tree size (total number of nodes in the tree used for generating the interactions).}
\label{fig:xyplot_treesize}
\end{figure}

@@

\begin{figure}[tb!]
\centering
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
levels(treespecs.long2$truetree.ind)[levels(treespecs.long2$truetree.ind)=="GLM tree"] <- "LM tree"
levels(treespecs.long2$truetree.ind)[levels(treespecs.long2$truetree.ind)=="GLMM tree"] <- "LMM tree"
aggdata.acc <- aggregate(formula = truetree.values ~ truetree.ind + N + sigmab + corUb, FUN = mean, data = treespecs.long2)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups = truetree.ind, type = "b", 
ylab = "tree accuracy", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), 
auto.key = list(space = "top", columns = 2, title = " ", cex.title = 1, lines = TRUE, points = TRUE)))
@
\caption{\small Tree accuracy of LM and LMM trees in simulated datasets. The $y$-axes represent the proportion of datasets in which the true tree was accurately recovered; $x$-axes represent the population standard deviation of the normal distribution from which the cluster-specific intercepts were drawn; rows represent dependence between the random intercepts ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[tb!]
\centering
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLM tree"] <- "LM tree"
levels(treespecs.long$correlation.ind)[levels(treespecs.long$correlation.ind)=="GLMM tree"] <- "LMM tree"
aggdata.cor <- aggregate(formula = correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN = mean, data = treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups = correlation.ind, 
type = "b", ylab = "correlation", xlab = expression(sigma[b]), par.settings = standard.theme("pdf", color = FALSE), 
auto.key = list(space = "top", columns = 3, title = " ", cex.title = 1, lines = TRUE, points = TRUE), ylim = c(0.3, 1.1)))
@
\caption{\small Average predictive accuracy of LM and LMM trees in simulated datasets. The $y$-axes represent the average correlation between the true and predicted treatment-effect differences; $x$-axes represent the population standard deviation of the normal distribution from which the cluster-specific intercepts were drawn; rows represent absolute treatment-effect differences in subgroups with treatment-effect differences; columns represent sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

Compared to LMMs with pre-specified interactions, LMM trees provided somewhat better accuracy, on average. LMM trees performed much better than LMMs when interactions were at least partly piecewise. However, the performance of LMM trees deteriorated when the interactions were purely continuous functions of the predictor variables, in which case LMMs with pre-specified interactions performed very well. The performance of LMMs deteriorated with a larger number of predictor variables, whereas the performance of LMM trees was not affected by this, confirming our expectation that LMM trees are better suited for exploration than LMMs.

@@

\begin{appendix}

\section{Glossary}

\subsection{Abbreviations}
